---
title: "Asteriod Data Analysis"
author: "Annika Lin, Hannah Norman, and Madeline Pfister"
date: "2023-05-02"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages and Functions

```{r}
library(caret)
library(mlbench)
library(rattle) 
library(randomForest)
library(ROCR)
library(glmnet)
library(car)
library(ResourceSelection)
```

## Predictive Performance Stats

We created a function in which the output is a table summarizing all of the predictive performance stats. This function was used throughout the analysis.
```{r}
get_stats <- function(CM) {
  TP <- CM[2,2]
  FP <- CM[2,1]
  TN <- CM[1,1]
  FN <- CM[1,2]
  
  acc <- (TP+TN) / (TP+TN+FN+FP)
  err <- (FP+FN) / (TP+TN+FN+FP)
  pre <- (TP) / (TP+FP)
  sen <- (TP) / (TP+FN)
  spe <- (TN) / (TN+FP)
  fme <- (2*pre*sen) / (pre+sen)
  mcc_denom <- sqrt(TP+FP)*sqrt(TP+FN)*sqrt(TN+FP)*sqrt(TN+FN)
  mcc <- (TP*TN - FP*FN) / mcc_denom
  
  name <- c("accuracy", "error rate", "precision", "sensitivity", "specificity", 
            "F-measure", "Matthew's CC")
  value <- c(acc, err, pre, sen, spe, fme, mcc)
  stats <- data.frame(name, value)
  
  return (stats)
}
```

# Data Cleaning and Initital Analysis

We began by choosing one date to focus on for our data set. The data we chose was April 6, 2017, so we filtered our data to only include observations from that date. 

```{r}
#import dataset filtered for '2017-04-06'
nasa_outlier <- read.csv("nasa_outlier.csv")
```

## Analysis of Predictors

We did a simple numerical summary and box plots to determine if there are any large outliers in the data set.
```{r}
# numerical summary + box plots
summary(nasa_outlier)
par(mfrow=c(2,4))
boxplot(nasa_outlier$Est.Dia.in.KM.range, main="Est.Dia.in.KM.range")
boxplot(nasa_outlier$Relative.Velocity, main="Relative.Velocity")
boxplot(nasa_outlier$Minimum.Orbit.Intersection, main="Minimum.Orbit.Intersection")
boxplot(nasa_outlier$Inclination, main="Inclination")
boxplot(nasa_outlier$Orbital.Period, main="Orbital.Period")
boxplot(nasa_outlier$Aphelion.Dist, main="Aphelion.Dist")
boxplot(nasa_outlier$Mean.Motion, main="Mean.Motion")
# to find row number of outlier observations
nasa_outlier[which.max(nasa_outlier$Est.Dia.in.KM.range),]
```

In an effort to identify extreme outliers, we selected 7 predictors whose numerical summaries suggested the existence of extreme outliers. Observation 695 was the only outlier were deemed extreme enough to remove. 

```{r}
#remove outlier
nasa <- nasa_outlier[-695,]
```

## Create a Proportion Table

We wanted to create a proportion table to get an understanding of how many asteroids were classified as hazardous and how many were classified as non-hazardous before beginning our analysis.
```{r}
prop.hazardous <- prop.table(table(nasa$Hazardous))
prop.hazardous
```

We found that the majority of our asteroids (85.9%) were classified as non-hazardous. All of our observations were classified as either hazardous or non-hazardous. This is also shown by the visualizations below.
```{r}
# pie chart
count.hazardous <- table(nasa$Hazardous)
lbls <- paste(levels(as.factor(nasa$Hazardous)), ": ",
              round(prop.hazardous,3)*100, "%", sep="")
pie(count.hazardous, labels=lbls, col=c("salmon", "palegreen"))
# bar plot
barplot(prop.hazardous, xlab="Hazardous", ylab="Proportion", ylim=c(0, 1.0),
        col=c("salmon", "palegreen"))
```
\newpage

# Data Analysis
## Lasso’s Penalized Regression

Create the design matrix.
```{r}
X = model.matrix(Hazardous ~ ., data=nasa)
Y = as.numeric(nasa$Hazardous=="True")
```

Conduct the cross-validation.
```{r}
set.seed(1)
cvfit = cv.glmnet(x=X[,-1], y=Y, family="binomial", type.measure="auc")
cvfit
plot(cvfit)
```

Determine which variables were selected using lambda.1se. These are the variables with which the lasso models will be built.
```{r}
sel.vars <- which(coef(cvfit, s=cvfit$lambda.1se)!=0)[-1]-1
sel.names <- colnames(nasa)[sel.vars]
sel.names
```

We built our initial lasso model using the variables selected.
```{r}
#fit a lasso model using the selected variables
fit.lasso <- glm(as.factor(Hazardous) ~ Absolute.Magnitude + Est.Dia.in.KM.min. +
                   Orbit.Uncertainity + Minimum.Orbit.Intersection + 
                   Mean.Motion, family="binomial", data=nasa)
summary(fit.lasso)
```

The model assumptions for independence, normality, and no influential points are satisfied by the data. However, we also need to check for multicollinearity to ensure that all model assumptions are satisfied.
```{r}
#address the model assumptions of the lasso model -- multicollinearity
vif(fit.lasso)
```

Both Absolute.Magnitude and Est.Dia.in.KM.min. have a VIF>5 indicating multicollinearity. We decided to initially adjust our model by removing Absolute.Magnitude as it has a larger VIF.
```{r}
#adjust the model based on multicollinearlity issues
#remove Absolute.Magnitude
fit.lasso2 <- glm(as.factor(Hazardous) ~ Est.Dia.in.KM.min. + Orbit.Uncertainity + 
                   Minimum.Orbit.Intersection + Mean.Motion,
                family="binomial", data=nasa)
summary(fit.lasso2)
```

```{r}
#assess multicollinearity of the adjusted model
vif(fit.lasso2)
```

Removing Absolute.Magnitude fixed the multicollinearity issues. All assumptions are satisfied for the adjusted model. We then tested how well the model fit the data using the Hosmer and Lemeshow goodness of fit test.

\begin{center}
$H_{0}$: the model fits the data well

$H_{1}$: the model does not fit the data well

$\alpha = 0.05$
\end{center}

```{r}
#test the goodness of fit of the adjusted model
hoslem.test(fit.lasso2$y, fit.lasso2$fitted.values)
```

There is statistically sufficient evidence (p = 0.436, df = 8) to conclude that the model fits the data well.

We then replaced Est.Dia.in.KM.min. with Absolute.Magnitude (adjusting for multicollinearity) to see if this model would be a better fit for the data.
```{r}
#adjust the model based on multicollinearlity issues
#remove Est.Dia.in.KM.min
fit.lasso3 <- glm(as.factor(Hazardous) ~ Absolute.Magnitude + Orbit.Uncertainity + 
                   Minimum.Orbit.Intersection + Mean.Motion,
                family="binomial", data=nasa)
summary(fit.lasso3)
```

```{r}
#assess multicollinearity of the adjusted model
vif(fit.lasso3)
```

There are no multicollinearity issues in the adjusted model.

Hosmer and Lemeshow goodness of fit test:

\begin{center}
$H_{0}$: the model fits the data well

$H_{1}$: the model does not fit the data well

$\alpha = 0.05$
\end{center}

```{r}
#test the goodness of fit of the adjusted model
hoslem.test(fit.lasso3$y, fit.lasso3$fitted.values)
```

There is statistically sufficient evidence (p = 0.5285, df = 8) to conclude that the model fits the data well.

Fit.lasso3 is the best fit of the lasso model as the Hosmer and Lemeshow goodness of fit (GOF) test results in a higher p-value for fit.lasso3 than fit.lasso2. This is the model in which we base the cross-validated prediction off of.
```{r}
#create the cross-validated model using the selected variables
set.seed(1)
fit.cv <- train(as.factor(Hazardous) ~ Absolute.Magnitude + Orbit.Uncertainity + 
                   Minimum.Orbit.Intersection + Mean.Motion , 
                   method = "glm", family = "binomial", 
                   trControl = trainControl(method="cv", number=5, 
                   savePredictions = TRUE, classProbs = TRUE),data=nasa)
fit.cv
#determine the final model
summary(fit.cv$finalModel)
```

Verify that there are no multicollinearity issues in the final model and that the model fits the data well.
```{r}
#asses the mulitcollinearity of the final model
vif(fit.cv$finalModel)
```

\begin{center}
$H_{0}$: the model fits the data well

$H_{1}$: the model does not fit the data well

$\alpha = 0.05$
\end{center}
```{r}
#goodness of fit
hoslem.test(fit.cv$finalModel$y, fit.cv$finalModel$fitted.values)
```

We then assessed the predictive performance of the model by plotting the ROC curve and finding the area under the curve.
```{r}
#plot the ROC curve
pihatcv <- fit.cv$pred
predcv <- prediction(pihatcv$True, pihatcv$obs)
perfcv <- performance(predcv, "tpr", "fpr")
plot(perfcv)
#find the area under the ROC curve
auccv <- performance(predcv, "auc")@y.values
auccv
```

Determine the importance of each of the predictors is used in the regression.
```{r}
plot(varImp(fit.cv), top=4)
```

Analyze the data analysis by creating a confusion matrix and generating performance statistics.
```{r}
confusionMatrix(pihatcv$pred, pihatcv$obs, positive="True") 
# Confusion matrix
conf.lasso <- table(pihatcv$pred, pihatcv$obs) 
conf.lasso
```

```{r}
lasso.stats <- get_stats(conf.lasso)
lasso.stats 
```

\newpage

## k-Nearest Neighbor

The kNN makes no assumptions and has one parameter to specify (k). We normalized our data before performing the analysis.
```{r}
# function to normalize data
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x))) }
arr.norm <- apply(nasa[,-21], 2, normalize)
arr.norm <- data.frame(arr.norm, nasa$Hazardous)
colnames(arr.norm)[colnames(arr.norm) == "nasa.Hazardous"] ="Hazardous"
```

We used the 5-fold cross validated model to choose k, and we found that 5 neighbors are used in the final model.
```{r}
# 5-fold CV to choose k
set.seed(1)
arr.norm$Hazardous <- as.factor(arr.norm$Hazardous)
fit.knn <- train(Hazardous ~ .,
  method = "knn",
  tuneGrid = expand.grid(k = 1:21),
  trControl = trainControl(method="cv", number=5, savePredictions = TRUE, classProbs = TRUE),
  metric = "Accuracy",
  data = arr.norm)
fit.knn
```

We then assessed the predictive performance of the model by plotting the ROC curve and finding the area under the curve, plotting the error rate, and plotting the most important variables.
```{r}
#plot the ROC curve
pihatcv.knn <- fit.knn$pred[fit.knn$pred$k == 5,]
pred <- prediction(pihatcv.knn$True, pihatcv.knn$obs)
perf <- performance(pred, "tpr", "fpr")
plot(perf)
# Area under ROC curve (AUC) = concordance index
auc.perf = performance(pred, "auc")
knn_auc <- auc.perf@y.values
knn_auc
```
The 10 most important variables in the k-Nearest neighbor models we built are:
```{r}
plot(varImp(fit.knn), main="kNN variable importance")
varImp(fit.knn)
```

Analyze the data by creating a confusion matrix and generating performance statistics.
```{r}
confusionMatrix(pihatcv.knn$pred, pihatcv.knn$obs, positive="True")
# Confusion matrix
conf.knn <- table(pihatcv.knn$pred, pihatcv.knn$obs) 
conf.knn
```

```{r}
knn.stats <- get_stats(conf.knn)
knn.stats 
```

\newpage

## Classification Tree Analysis

There are no model assumptions to be satisfied for decision trees. We are fitting a classification tree as opposed to a regression tree because the response variable is categorical and binary.
```{r}
set.seed(1)
# rpart
nasa.CVrpart <- train(Hazardous ~ ., data=nasa,
                      method="rpart",
                      tuneGrid = expand.grid(cp=seq(0.005, 0.05, length=10)),
                      trControl=trainControl(method="cv", number=5,
                                             savePredictions=TRUE,
                                             classProbs=TRUE,
                                             selectionFunction = "oneSE"))
nasa.CVrpart
# print tree
fancyRpartPlot(nasa.CVrpart$finalModel)
```

If the plain rpart method is utilized, it selects a cp (complexity parameter) value of 0.05 to maximize accuracy for the final model. Note that the accuracy and kappa values are identical for each of the cp values tried by the model, so any choice within that range should produce comparable results.

We then assessed the predictive performance by plotting the ROC curve and finding the area under the curve.
```{r}
pihatcv.rpart <- nasa.CVrpart$pred[nasa.CVrpart$pred$cp == 0.05,]
predcv.rpart <- prediction(pihatcv.rpart$True, pihatcv.rpart$obs)
perfcv.rpart <- performance(predcv.rpart, "tpr", "fpr")
plot(perfcv.rpart)
aucCV.rpart <- performance(predcv.rpart, "auc")@y.values 
aucCV.rpart
```

The 10 most important variables in the model are:
```{r}
# variable importance
varImp(nasa.CVrpart)
plot(varImp(nasa.CVrpart))
```

Analyze the data by creating a confusion matrix and generating performance statistics.
```{r}
confusionMatrix(pihatcv.rpart$pred, pihatcv.rpart$obs, positive="True")
# Confusion matrix
conf.rpart <- table(pihatcv.rpart$pred, pihatcv.rpart$obs) 
conf.rpart
```

```{r}
rpart.stats <- get_stats(conf.rpart)
rpart.stats
```

\newpage

## Random Forest Analysis

There are no model assumptions to be satisfied for random forest analysis.
```{r}
set.seed(1)
nasa.rf <- randomForest(as.factor(Hazardous) ~ ., data=nasa)
nasa.rf
```

```{r}
set.seed(1)
nasa.CVrf <- train(Hazardous ~ ., data=nasa,
                   method="rf",
                   trControl=trainControl(method="cv", number=5,
                                          savePredictions=TRUE,
                                          classProbs=TRUE))
nasa.CVrf
```

The final model selects an mtry tuning parameter value of 20 in order to maximize accuracy.

We then assessed the model predictive performance by plotting the ROC curve and finding the model under the curve.
```{r}
pihatcv.rf <- nasa.CVrf$pred[nasa.CVrf$pred$mtry == 20,]
predcv.rf <- prediction(pihatcv.rf$True, pihatcv.rf$obs)
perfcv.rf <- performance(predcv.rf, "tpr", "fpr")
plot(perfcv.rf)
aucCV.rf <- performance(predcv.rf, "auc")@y.values 
aucCV.rf
```

The 10 most important variables is the random forest model are:
```{r}
varImp(nasa.CVrf)
plot(varImp(nasa.CVrf))
```

Analyze the data by creating a confusion matrix and generating performance statistics.
```{r}
confusionMatrix(pihatcv.rf$pred, pihatcv.rf$obs, positive="True")
# Confusion matrix
conf.rf <- table(pihatcv.rf$obs, pihatcv.rf$pred)
conf.rf
```

```{r}
rf.stats <- get_stats(conf.rf)
rf.stats
```

\newpage

# Model Comparisons

## Variable Importance Comparison
```{r}
imp.knn <- rownames(varImp(fit.knn)$importance)
sel.knn <- imp.knn[order(varImp(fit.knn)$importance[,1], decreasing=T)][1:10]
imp.rp <- rownames(varImp(nasa.CVrpart)$importance)
sel.rp <- imp.rp[order(varImp(nasa.CVrpart)$importance[,1], decreasing=T)][1:10]
imp.rf <- rownames(varImp(nasa.CVrf)$importance)
sel.rf <- imp.rf[order(varImp(nasa.CVrf)$importance[,1], decreasing=T)][1:10]
intersect(sel.names, intersect(sel.knn, intersect(sel.rp, sel.rf)))
```

## Predicitive Performance Stat Comparison
```{r}
df_merge <- merge(lasso.stats,knn.stats,by="name")
colnames(df_merge)[colnames(df_merge) == "value.x"] ="Lasso"
colnames(df_merge)[colnames(df_merge) == "value.y"] ="kNN"
df_merge <- merge(df_merge,rpart.stats,by="name")
df_merge <- merge(df_merge,rf.stats,by="name")
colnames(df_merge)[colnames(df_merge) == "value.x"] ="CT"
colnames(df_merge)[colnames(df_merge) == "value.y"] ="RF"
df_merge
```



